---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slopes

<!-- badges: start -->
![R CMD Check via {tic}](https://github.com/ITSLeeds/slopes/workflows/R%20CMD%20Check%20via%20%7Btic%7D/badge.svg)
<!-- badges: end -->

The goal of slopes is to enable fast, accurate and user friendly calculation longitudinal steepness of linear features such as roads and rivers, based on commonly available input datasets such as road geometries and digital elevation model (DEM) datasets.

## Installation

<!-- You can install the released version of slopes from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("slopes") -->
<!-- ``` -->

Install the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("itsleeds/slopes")
```

## Usage

Load the package in the usual way:

```{r}
library(slopes)
```

We will also load the `sf` library:

```{r}
library(sf)
```


<!-- We will also use the `sf` package for representing road segments: -->

<!-- ```{r} -->
<!-- library(sf) -->
<!-- ``` -->

The minimum data requirements for using the package are elevation points, either as a vector, a matrix or as a digital elevation model (DEM) encoded as a raster dataset.
Typically you will also have a geographic object representing the roads or similar features.
These two types of input data are represented in the code output and plot below.

```{r dem-lisbon}
# A raster dataset included in the package:
class(dem_lisbon_raster) # digital elevation model
summary(raster::values(dem_lisbon_raster)) # heights range from 0 to ~100m
raster::plot(dem_lisbon_raster)

# A vector dataset included in the package:
class(lisbon_road_segments)
plot(sf::st_geometry(lisbon_road_segments), add = TRUE)
```

Calculate the average gradient of each road segment as follows:

```{r}
lisbon_road_segments$slope = slope_raster(lisbon_road_segments, e = dem_lisbon_raster)
summary(lisbon_road_segments$slope)
```

This created a new column, `slope` that represents the average, distance weighted slope associated with each road segment.
The units represent the percentage incline, that is the change in elevation divided by distance.
The summary of the result tells us that the average gradient of slopes in the example data is just over 5%.
This result is equivalent to that returned by ESRI's `Slope_3d()` in the [3D Analyst extension](https://desktop.arcgis.com/en/arcmap/10.3/tools/3d-analyst-toolbox/slope.htm), with a correlation between the ArcMap implementation and our implementation of more than 0.95 on our test datast (we find higher correlations on larger datasets):

```{r}
cor(
  lisbon_road_segments$slope,    # slopes calculates by the slopes package
  lisbon_road_segments$Avg_Slope # slopes calculated by ArcMap's 3D Analyst extension
)
```

We can now visualise the slopes calculated by the `slopes` package as follows:

```{r slope-vis}
raster::plot(dem_lisbon_raster)
plot(lisbon_road_segments["slope"], add = TRUE, lwd = 5)
# mapview::mapview(lisbon_road_segments["slope"], map.types = "Esri.WorldStreetMap")
```

Imagine that we want to go from Santa Catarina to the East of the map to the Castelo de Sao Jorge to the West of the map:

```{r route}
mapview::mapview(lisbon_route)
```


We can convert the `lisbon_route` object into a 3d linestring object as follows:

```{r, warning=FALSE, echo=FALSE, eval=FALSE}
sln = stplanr::SpatialLinesNetwork(lisbon_road_segments)
points = sf::st_as_sf(crs = 4326, coords = c("X1", "X2"), data.frame(rbind(
  stplanr::geo_code("Santa Catarina, Lisbon"),
  stplanr::geo_code("Castelo, Lisbon")
)))
points_projected = sf::st_transform(points, sf::st_crs(lisbon_road_segments))
coords = sf::st_coordinates(points_projected)
nodes = stplanr::find_network_nodes(sln, coords[, 1], coords[, 2])
lisbon_route = stplanr::sum_network_routes(sln = sln, start = nodes[1], end = nodes[2])
mapview::mapview(lisbon_route) +
  mapview::mapview(lisbon_road_segments["slope"])
lisbon_route_3d = slope_3d(lisbon_route, dem_lisbon_raster)
usethis::use_data(lisbon_route, overwrite = TRUE)
usethis::use_data(lisbon_route_3d, overwrite = TRUE)
```

```{r}
lisbon_route_3d = slope_3d(lisbon_route, dem_lisbon_raster)
```

We can now visualise the elevation profile of the route as follows:

```{r plot_slope}
plot_slope(lisbon_route_3d)
```

# Performance

For this benchmark we will download the following small (< 100 kB) `.tif` file:

```{r}
u = "https://github.com/ITSLeeds/slopes/releases/download/0.0.0/dem_lisbon.tif"
if(!file.exists("dem_lisbon.tif")) download.file(u, "dem_lisbon.tif")
```

A benchmark can reveal how many route gradients can be calculated per second:

```{r}
e = dem_lisbon_raster
r = lisbon_road_segments
et = terra::rast("dem_lisbon.tif")
res = bench::mark(check = FALSE,
  slope_raster = slope_raster(r, e, terra = FALSE),
  slope_terra1 = slope_raster(r, e, terra = TRUE),
  slope_terra2 = slope_raster(r, et, terra = TRUE)
)
res
```

That is approximately

```{r}
round(res$`itr/sec` * nrow(r))
```

routes per second using the `raster` and `terra` (the default if installed, using `RasterLayer` and native `SpatRaster` objects) packages to extract elevation estimates from the raster datasets, respectively.

The message: use the `terra` package to read-in DEM data for slope extraction if speed is important.

To go faster, you can chose the `simple` method to gain some speed at the expense of accuracy:

```{r}
e = dem_lisbon_raster
r = lisbon_road_segments
res = bench::mark(check = FALSE,
  bilinear1 = slope_raster(r, e, terra = TRUE),
  bilinear2 = slope_raster(r, et, terra = TRUE),
  simple1 = slope_raster(r, e, method = "simple", terra = TRUE),
  simple2 = slope_raster(r, et, method = "simple", terra = TRUE)
)
# ?bench::mark
res
```

The equivalent benchmark with the `raster` package is as follows:

```{r}
e = dem_lisbon_raster
r = lisbon_road_segments
res = bench::mark(check = FALSE,
  bilinear = slope_raster(r, e, terra = FALSE),
  simple = slope_raster(r, e, method = "simple", terra = FALSE)
)
# ?bench::mark
res
```

```{r, echo=FALSE}
file.remove("dem_lisbon.tif")
```


<!-- That is sufficient for our needs but we plan to speed-up the calculation, e.g. using the new `terra` package, as outlined this [thread](https://github.com/rspatial/terra/issues/29#issuecomment-619444555). -->

