---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slopes

<!-- badges: start -->
![R CMD Check via {tic}](https://github.com/ITSLeeds/slopes/workflows/R%20CMD%20Check%20via%20%7Btic%7D/badge.svg)
<!-- badges: end -->

The goal of slopes is to enable rapid and accurate calculation slope steepness of linear features such as roads, based on commonly available input datasets such as road geometries and digital elevation model (DEM) datasets.

## Installation

<!-- You can install the released version of slopes from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("slopes") -->
<!-- ``` -->

Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("itsleeds/slopes")
```

## Usage

Load the package in the usual way:

```{r}
library(slopes)
```

<!-- We will also use the `sf` package for representing road segments: -->

<!-- ```{r} -->
<!-- library(sf) -->
<!-- ``` -->

The minimum data requirements for using the package are elevation points, either as a vector, a matrix or as a digital elevation model (DEM) encoded as a raster dataset.
Typically you will also have a geographic object representing the roads or similar features.
These two types of input data are represented in the code output and plot below.

```{r dem-lisbon}
# A raster dataset included in the package:
class(dem_lisbon_raster) # digital elevation model
summary(raster::values(dem_lisbon_raster)) # heights range from 0 to ~100m
raster::plot(dem_lisbon_raster)

# A vector dataset included in the package:
class(lisbon_road_segments)
plot(sf::st_geometry(lisbon_road_segments), add = TRUE)
```

Calculate the average gradient of each road segment as follows:

```{r}
lisbon_road_segments$slope = slope_raster(lisbon_road_segments, e = dem_lisbon_raster)
summary(lisbon_road_segments$slope)
```

This created a new column, `slope` that represents the average, distance weighted slope associated with each road segment.
The units represent the percentage incline, that is the change in elevation divided by distance.
The summary of the result tells us that the average gradient of slopes in the example data is just over 5%.
This result is equivalent to that returned by ESRI's `Slope_3d()` in the [3D Analyst extension](https://desktop.arcgis.com/en/arcmap/10.3/tools/3d-analyst-toolbox/slope.htm), with a correlation between the ArcMap implementation and our implementation of more than 0.95 on our test datast (we find higher correlations on larger datasets):

```{r}
cor(
  lisbon_road_segments$slope,    # slopes calculates by the slopes package
  lisbon_road_segments$Avg_Slope # slopes calculated by ArcMap's 3D Analyst extension
)
```

We can now visualise the slopes calculated by the `slopes` package as follows:

```{r slope-vis}
raster::plot(dem_lisbon_raster)
plot(lisbon_road_segments["slope"], add = TRUE, lwd = 5)
```

# Performance

A benchmark can reveal how many gradients are calculated per second:

```{r}
e = dem_lisbon_raster
r = lisbon_road_segments[1:100, ]
bench::mark(
  slope_raster(r, e)
)
```


