---
title: "Debugging issues in the slopes package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Debugging issues in the slopes package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slopes)
```

The aim of this vignette is to document issues that have been identified and (hopefully, eventually) fixed.
It does not run by default.
To run it, change `FALSE` to `TRUE` in the code chunk below.

```{r}
knitr::opts_chunk$set(eval = FALSE)
```


## Error in weighted mean (#12)

This issue showed that the function `slope_raster()` errors in the final line:

```{r}
library(sf)
library(raster)
library(slopes)

#import datasets
network_original = st_read("https://github.com/U-Shift/Declives-RedeViaria/blob/main/shapefiles/RedeViariaLisboa_dadosabertos.gpkg?raw=true")
network = st_zm(network_original, drop = T) # make sure it has no Z values stored
network
u = "https://github.com/U-Shift/Declives-RedeViaria/blob/main/raster/LisboaNASA_clip.tif?raw=true"
download.file(u, "LisboaNASA_clip.tif")
dem = raster("LisboaNASA_clip.tif")
dem

# do they overlap?
raster::plot(dem)
plot(sf::st_geometry(network), add = TRUE)
# why this error?
network$slopeNASA = slope_raster(network, e = dem)
# In addition: Warning messages:
# 1: In max(d, na.rm = TRUE) :
#   no non-missing arguments to max; returning -Inf
# 2: In max(d, na.rm = TRUE) :
#   no non-missing arguments to max; returning -Inf
```

Does it error on a subset of the network?

```{r}
network_subset = network[1:9, ]
network$slopesNASA = slope_raster(network_subset, e = dem) # also fails with the same error
dem_subset = crop(dem, network)
network_subset$slopesNASA = slope_raster(network_subset, e = dem_subset) # also fails with the same error
unique(st_geometry_type(network_subset))
unique(st_geometry_type(lisbon_road_segment))
mapview::mapview(network_subset)
network_subset_ls = sf::st_cast(network_subset, to = "LINESTRING")
mapview::mapview(network_subset_ls)
network_subset_ls$slopesNASA = slope_raster(network_subset_ls, e = dem_subset) 
```

Testing on the full network:


```{r}
# install latest version
remotes::install_github("itsleeds/slopes")
library(sf)
library(raster)
library(slopes)

#import datasets
network_original = st_read("https://github.com/U-Shift/Declives-RedeViaria/blob/main/shapefiles/RedeViariaLisboa_dadosabertos.gpkg?raw=true")
network = st_zm(network_original, drop = T) # make sure it has no Z values stored
network
u = "https://github.com/U-Shift/Declives-RedeViaria/blob/main/raster/LisboaNASA_clip.tif?raw=true"
download.file(u, "LisboaNASA_clip.tif")
dem = raster("LisboaNASA_clip.tif")
dem

# do they overlap?
raster::plot(dem)
plot(sf::st_geometry(network), add = TRUE)
# why this error?
network$slopeNASA = slope_raster(network, e = dem)
network_ls = sf::st_cast(network, "LINESTRING")
network_ls$slopeNASA = slope_raster(network_ls, e = dem)
plot(network_ls["slopeNASA"], lwd = 5)
```

# Issues with terra

```{r}
library(slopes)
class(lisbon_road_segments)
raster::plot(dem_lisbon_raster)
plot(sf::st_geometry(lisbon_road_segments), add = TRUE)
demterra = terra::rast(dem_lisbon_raster)
plot(sf::st_geometry(lisbon_road_segments), add = TRUE)
lisbon_road_segments$sloperaster = slope_raster(lisbon_road_segments, e = dem_lisbon_raster)
library(terra)
plot(demterra)
lisbon_road_segments$slopeterra = slope_raster(lisbon_road_segments, e = demterra, terra = T)
summary(lisbon_road_segments$sloperaster)
summary(lisbon_road_segments$slopeterra)

# Looking at the code, this seems to be the culprit:

# res = as.numeric(terra::extract(e, m[, 1:2], method = method))
# vs
# res = as.numeric(raster::extract(e, m[, 1:2], method = method))
# lets try it...
method = "bilinear"
m = sf::st_coordinates(lisbon_road_segment)[1:2, ]
e = demterra
as.numeric(terra::extract(e, m[, 1:2], method = method))
as.numeric(terra::extract(dem_lisbon_raster, m[, 1:2], method = method))
res_terra = terra::extract(e, m[, 1:2], method = method)
class(res_terra)
res_terra
as.numeric(res_terra[, "r1"])
```


