---
title: "Benchmarking slopes calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking slopes calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(slopes)
library(bench)
```

**Note: benchmarks do not currently evaluate on actions for faster build times.**

# Performance

For this benchmark we will download the following small (< 100 kB) `.tif` file:

```{r}
u = "https://github.com/ITSLeeds/slopes/releases/download/0.0.0/dem_lisbon.tif"
f = basename(u)
if(!file.exists(f)) {
 download.file(u, f, mode = "wb") 
}
```

A benchmark can reveal how many route gradients can be calculated per second:

```{r, results='hide'}
e = dem_lisbon_raster
r = lisbon_road_network
et = terra::rast(f)
res = bench::mark(check = FALSE,
  slope_raster = slope_raster(r, e),
  slope_terra = slope_raster(r, et)
)
```

```{r}
res
```


That is approximately

```{r}
round(res$`itr/sec` * nrow(r))
```

routes per second using the `raster` and `terra` (the default if installed, using `RasterLayer` and native `SpatRaster` objects) packages to extract elevation estimates from the raster datasets, respectively.

The message: use the `terra` package to read-in DEM data for slope extraction if speed is important.

To go faster, you can chose the `simple` method to gain some speed at the expense of accuracy:

```{r, results='hide'}
e = dem_lisbon_raster
r = lisbon_road_network
res = bench::mark(check = FALSE,
  bilinear1 = slope_raster(r, e),
  bilinear2 = slope_raster(r, et),
  simple1 = slope_raster(r, e, method = "simple"),
  simple2 = slope_raster(r, et, method = "simple")
)
```

```{r}
res
```


```{r, include=FALSE}
file.remove("dem_lisbon.tif")
```
